---
title: "measles"
subtitle: "A vignette."
author: "Clemens Ager"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
link-citations: yes
---

Note - I don't know the us american system. So I speculate on some correlations. 
Mostly I suppose they don't do vaccination against the parents will, but rather I'm seeing honest typos in the data.
Not too motivated to organized secondary sources.

## Setup
```{r}
getwd()
```

```{r setup}
library(ggplot2)
library(dplyr)
library(tidyr)
library(pander)

measles <- readRDS("data/measles.RDS")

cmeasles  <- measles %>% filter(!suspicious)
```

## checking data

I notice two annoyances:

  1. In some columns `-1` is used instead of `NA`
  2. Religious refusal is boolean ?!?
  2. Lots of missing values.
  3. The ratios of vaccinations and students refusing don't add up.
  4. In one school `160%` refused for other reasons.

### weirdness
```{r, fig.cap="correspondance mmr with xmed"}
ptm <- proc.time()
ggplot(measles, mapping=aes(x=mmr, y=replace_na(xmed,0))) + 
  geom_point(alpha = 0.2) 
proc.time()-ptm
```
```{r, fig.cap="correspondance mmr with xper"}
ggplot(measles, mapping=aes(x=mmr, y=replace_na(xper,0))) + 
  geom_point(alpha = 0.2) 
```

These are suspicious. There is a clearly a diagonal line visible. 
There should not be points above - either they withdrew the consent, or were wrongly reported vaccinated.
It seems many just report 100%. 
And the magic school where 169% rejected - on the figure this looks like a reasonable typo - 16.9% may fit. But still the `mmr` is 92.3% - way too high.

```{r weird entries}
susps <- measles %>% filter(suspicious) %>% select(-lng,-lat)
```
In total there are `r nrow(susps)` suspicious entries - that's not much compared out of `r nrow(measles)`.


### Imputation

I didn't want to exclude all the `NA`s.

#### xmed and xper

seem simple.  They are `NA` for `100%` schools. 
So in most cases the `NA` just may be `0`.

#### xrel

is harder as it is not numerically reported. 
But mmr and xmed seem to be reliable - so I decided to have all *unexplained* non-vaccing together.  Of course this will group religious with all where no explanation was required.

#### School type
School type seemets interesting, but has lots of `NA`s.  
Seems the reason is most states don't report these. 
Not much I can do there.
```{r mmr unreported}
pander(measles %>% group_by(type, state) %>% summarise(n=n()) %>%
       spread(type,n), caption = "types of school vs. states",
     split.tables = Inf)
```

#### Enroll

Number of students would be useful for a weighted average.

```{r mmr unreported}
pander(measles %>% group_by( state) %>% 
       summarise(median=median(enroll, na.rm = T),
		 mad = mad(enroll, na.rm = T)),
     caption = "types of school vs. states",
     split.tables = Inf)
```

The overall median is `r median(measles$enroll, na.rm = T)` and 
mad `r mad(measles$enroll, na.rm = T)`.
I chose 100 as guess for the undocumented ones so that there would be a nice log10.


```{r, fig.cap="distribution mmr(measles plus vac)"}
ggplot(cmeasles, mapping=aes(x=mmr)) + 
  geom_histogram(binwidth=5) +
  scale_y_log10() + annotation_logticks(sides=c("l")) +
  theme_light()
```

```{r, fig.cap="distribution of overall vac"}
ggplot(cmeasles, mapping=aes(x=overall)) + 
  geom_histogram(binwidth=5) +
  scale_y_log10() + annotation_logticks(sides=c("l"))
```

```{r, fig.cap="correspondance mmr with vac"}
ggplot(measles, mapping=aes(x=overall, y=mmr)) + 
  geom_point(alpha = 0.2) 
```
A few odd ones where overall vaccination is lower than mmr. But since I don't use overall later I will ignore this.

xrel is bool!
```{r, fig.cap="distribution of mmr in schools with religious exclusions"}
ggplot(cmeasles, mapping=aes(x=mmr)) + 
  geom_histogram(binwidth=5) +
  scale_y_log10() + annotation_logticks(sides=c("l"))
```

It should be mmr + xmed < 100!

```{r, fig.cap="distribution of mmr in schools with xmed"}
ggplot(cmeasles, mapping=aes(y=mmr, x=xmed)) + 
  geom_point()
cmeasles %>% filter(xmed > 50)
```
```{r, fig.cap="distribution of mmr in schools with xper"}
ggplot(cmeasles, mapping=aes(y=mmr, x=xper)) + 
  geom_point()
cmeasles %>% filter(xmed > 50)
```


```{r, fig.cap="distribution of mmr in schools with either exemption"}
ggplot(cmeasles, mapping=aes(x=xper)) + 
  geom_histogram(binwidth=5) +
  scale_y_log10() + annotation_logticks(sides=c("l"))

ggplot(cmeasles, mapping=aes(x=xper,y=mmr)) + 
  geom_point()
```


columns:
  - state: US state
  - enroll: number of students
  - mmr: vaccination ratio (percent)

Note: 
  - has only one sample per school.
  - only 15 states

```{r}
msum <- measles %>% 
  filter(!is.na(enroll), mmr>=0) %>%
  group_by(state, year) %>%
  summarize(allpup=sum(enroll), nummmr=(sum(enroll*mmr/100))) %>%
  ungroup() %>%
  mutate(ratio = nummmr/allpup)

ggplot(msum, mapping = aes(x=state, y=ratio)) + geom_point()
```

```{r kinds of school}
cmeasles %>% group_by(type) %>% summarise(n=n())%>% arrange(-n)
```
Lots of unlabelled. Private don't play much of a role.

```{r states}
cmeasles %>% group_by(state) %>% summarise(n=n(),enrolls = sum(enroll, na.rm=T)) %>% arrange(-n)
```
Everything is dwarfed by California and Illinois.


## Visualisierung:

### Idea ternary plot:

```{r}
library(ggtern)
cm2 <- cmeasles %>% 
  mutate(xmed = replace_na(xmed,0),
	 xother = 100 - mmr - xmed,
	 enroll = replace_na(enroll,100)) %>%
  mutate_at(vars(mmr, xmed, xother), ~./100)
cms <- sample_n(cm2,100)

my_breaks  <-  10^seq(2,6)
```

```{r each school}
ggtern(cm2, aes(mmr,xmed,xother)) +
  geom_point_swap(map = aes(fill = type), 
		  alpha = 0.5) + 
  scale_L_continuous("vaccinated") + 
  scale_R_continuous("refuse other") + 
  scale_T_continuous("refuse medical") + 
  theme_tropical()
  #theme_linedraw()
  #theme_bvbg()

```

```{r}
ptm <- proc.time()
#cms <- cm2 %>% filter(type == "Kindergarten")
ggtern(cm2, aes(mmr,xmed,xother)) +
  geom_tri_tern(bins = 30, aes(fill=..stat.., value=enroll), fun=function(z){(sum(z))}) +
  scale_L_continuous("vaccinated") + 
  Llab("measles vaccinated") +
  scale_R_continuous("other") + 
  Rlab("other refusal") + 
  scale_T_continuous("medical") + 
  Tlab("medical refusal") +
  scale_fill_gradient(name = "count", trans = "log",
		      breaks = my_breaks, labels = my_breaks,
		      na.value = "gray10") + 
  theme_matrix() +
  theme(legend.background = element_rect(fill = "gray10")) 

  # theme_tropical()
  #theme_linedraw()
  #theme_bvbg()
  # scale_fill_continuous(low="thistle2", high="darkred", 
  #                      guide="colorbar",na.value="white") +
proc.time()-ptm

```


